---
title: "Module 1: The Magic of One Function"
author: "NCME 2026 PGDCM Workshop"
output: html_document
---

# The Magic of One Function

**Objective:** Move from standard Q-matrices to full Bayesian estimation natively utilizing the `run_pgdcm_auto` pipeline.

## 1. Data Import

First, we will load our sample student responses and the theoretical Q-matrix representing our items.

```{r load-data, message=FALSE, warning=FALSE}
library(pgdcm)
library(dcmdata)

# Load the dataset (Student Responses) and Q-matrix (Item-Skill mapping)
X <- dtmr_data
Q <- dtmr_qmatrix
```

## 2. PGDCM Configuration

Traditional DCMs rely on rigid Q-matrices. We will convert this matrix into a Directed Acyclic Graph (DAG) for `pgdcm` and build our model configuration.

```{r make-graph}
# Convert the classical Q-matrix to a Directed Acyclic Graph (DAG)
# FILL IN THE BLANK: Use the appropriate pgdcm helper function
g_dcm <- ___(Q)

# Build the model configuration linking the behavioral data (X) to the graph
# FILL IN THE BLANK: Provide the graph object where requested
config_dcm <- build_model_config(graph = ___, data = X)
```

## 3. The Bayesian Engine

Now we pass the combined configuration object into our fully automated Bayesian estimation pipeline. 

```{r run-mcmc, eval=FALSE}
# We are reducing MCMC iterations here purely to save time in the workshop.
# FILL IN THE BLANK: Provide the config_dcm object
results_dcm <- run_pgdcm_auto(
  config = ___,
  prefix = "NCME_DCM",
  estimation_config = list(niter = 2500, nburnin = 500, chains = 2) 
)

# Save results for later use (Module 3 & 4)
saveRDS(results_dcm, "results_dcm.rds")
```

## 4. Review the Outputs
Once estimation finishes, let's look at the primary objects `pgdcm` returns.

```{r view-results}
# Look at the latent skill profiles (1 = Master, 0 = Non-Master) for a few students
head(results_dcm$skill_profiles)

# Look at the calibrated item parameters (e.g., slipping and guessing rates)
head(results_dcm$item_parameters)
```
