---
title: "Module 2: Visualizing Cognitive Architectures"
author: "NCME 2026 PGDCM Workshop"
output: html_document
---

# Visualizing Cognitive Architectures

**Objective:** Move beyond rigid Q-matrices to map complex learning progressions and mixed response rules visually using Cytoscape.

## 1. Connecting R to Cytoscape

Cytoscape is an incredible tool for mapping Bayesian networks perfectly. Ensure the Cytoscape app is physically open on your computer before running this block.

```{r setup-cytoscape, message=FALSE, warning=FALSE}
library(pgdcm)
library(igraph)
library(RCy3) 

# Verify Cytoscape is running
cytoscapePing()
```

## 2. Drawing the Foundational Code Path

Suppose we have a cognitive theory: `Skill_A` is a strict prerequisite for `Skill_B`. Let's lay this foundation in R first.

```{r base-graph}
# Create a basic foundational graph in R
# FILL IN THE BLANK: Define the directed edge from A to B.
g_custom <- make_graph(edges = c("___", "___"), directed = TRUE)

# Send the foundational graph to Cytoscape
createNetworkFromIgraph(g_custom, title = "Workshop: Learning Progression")
```

## 3. The Cytoscape Intervention (GUI Task)

🛑 **STOP: Move over to the Cytoscape Application** 🛑

**Your Tasks in Cytoscape:**
1. Add a new node called `Skill_C`. Draw a directed edge from `Skill_B` -> `Skill_C`.
2. Add an item node `Item_1` and connect it to `Skill_C`.
3. Select `Item_1` in the node table window and set its `compute_type` explicitly to `DINO`.

## 4. Pulling Back to R

We now pull the modified, highly customized architecture back down to R to prep for parameter estimation.

```{r pull-graph}
# Pull the modified architecture back into R
# FILL IN THE BLANK: Enter the title of the network we created above
g_modified <- createIgraphFromNetwork(network = "___")

# Save the visual network as a structured GraphML object 
write_graph(g_modified, file = "my_custom_model.graphml", format = "graphml")

cat("\nModule 2 Complete: Complex architecture saved and ready for pgdcm!\n")
```
