---
title: "Module 3: The Model Showdown & Diagnostics"
author: "NCME 2026 PGDCM Workshop"
output: html_document
---

# The Model Showdown & Diagnostics

**Objective:** Critically evaluate which cognitive architecture best fits empirical data using native Bayesian fit indices.

We'll start by loading the models we processed during the break from previous modules. 

```{r load-results, message=FALSE, warning=FALSE}
library(pgdcm)
library(MCMCvis)

# Load the results we generated in an earlier module
results_dcm <- readRDS("results_dcm.rds")
# results_custom <- readRDS("results_custom.rds") # Use this if you finished Module 2 runtime
```

## Section A: Relative Model Fit (WAIC)

The Widely Applicable Information Criterion (WAIC) penalizes models for unnecessary complexity. **Lower WAIC = Better Fit.**

Which of our cognitive architectures is the true "winner"? Let's find out effortlessly.

```{r waic-comparison}
# FILL IN THE BLANK: Extract the WAIC variable from the results object
print(paste("Standard DCM WAIC:", results_dcm$___))
# print(paste("Custom Architecture WAIC:", results_custom$___))
```

## Section B: MCMC Convergence Diagnostics

Because we are doing full Bayesian sampling, we must verify the MCMC chains reached their stationary distributions. We want **Gelman-Rubin $\hat{R}$ values close to 1.0** (ideally < 1.1).

```{r mcmc-diagnostics}
# Retrieve a numerical summary table specifically for the 'lambda' item parameters
# FILL IN THE BLANK: Pass the MCMC samples from the results object
res_summary <- MCMCsummary(___$samples, params = "lambda")
head(res_summary)
```

With one function, we can visually plot the posterior credible intervals for these parameters:

```{r plot-diagnostics, fig.height=6, fig.width=8}
# Visually plot the 'lambda' parameter distributions to look at the credible intervals
MCMCplot(results_dcm$samples, params = "lambda")
```
