---
title: "Module 4: Operational Scoring Engine"
author: "NCME 2026 PGDCM Workshop"
output: html_document
---

# Operational Scoring Engine

**Objective:** Transition from exploratory calibration to high-speed operational scoring of new students using locked priors.

## 1. Defining the Gold Standard

Calibration is heavy. Operational student scoring should be fast. We will take the item parameters estimated in Hour 1 and lock them in as "Gold Standard" assumptions.

```{r load-priors, message=FALSE, warning=FALSE}
library(pgdcm)
library(dcmdata)

# Identify the 'winning' parameters from earlier calibration
gold_standard_params <- read.csv("NCME_DCM_item_parameters.csv")
head(gold_standard_params)
```

## 2. Acquiring Operational Data

We simulate acquiring a brand NEW, unseen dataset of operational student responses. For this exercise, we'll pretend the second half of our dataset is brand new.

```{r config-operational}
# For this exercise, we'll pretend the second half of `dtmr_data` is brand new.
X_new <- dtmr_data[501:1000, ] 
Q <- dtmr_qmatrix

g_dcm <- QMatrix2iGraph(Q)

# Build the scoring configuration
# FILL IN THE BLANK: Map the graph to the NEW operational data
config_scoring <- build_model_config(graph = g_dcm, data = ___)
```

## 3. The Scoring Engine

By feeding `pgdcm` our prior parameter expectations with near-zero variance, the engine skips recalibrating items entirely and converges rapidly on predicting latent skills for the new students.

```{r run-scoring, eval=FALSE}
# Run `pgdcm` in purely operational scoring mode!
# FILL IN THE BLANK: You might need to adjust the structure based on pgdcm's `priors` argument layout.
results_scoring <- run_pgdcm_auto(
  config = config_scoring,
  prefix = "NCME_Scoring",
  # Note: Replace `priors_list` with the properly formatted list of locked parameter means and tiny standard deviations. 
  # priors = priors_list, 
  estimation_config = list(niter = 2000, nburnin = 500, chains = 2)
)
```

## 4. Diagnostics Delivered

Just like that, we successfully scored hundreds of new students using our validated engine natively!

```{r final-results, eval=FALSE}
# Output the profiles for the NEW students
head(results_scoring$skill_profiles)

cat("\nModule 4 Complete: Operational Scoring achieved!\n")
```
