---
title: "Scoring Cookbook: Operational Scoring & Cross-Validation"
format:
  html:
    toc: true
    number-sections: true
    html-math-method: katex
vignette: >
  %\VignetteIndexEntry{Scoring Cookbook: Operational Scoring & Cross-Validation}
  %\VignetteEngine{quarto::html}
---

## Introduction: Operational Scoring

::: callout-tip
## Learning Objectives

By the end of this cookbook, you will learn how to:

1.  Split a dataset into Calibration and Scoring samples to simulate a multi-year assessment cycle.
2.  Run a full Calibration phase using `build_model_config()` and `run_pgdcm_auto()`.
3.  Lock item parameters and transfer them to new examinees using `build_scoring_config()`.
4.  Evaluate model generalization via cross-validated accuracy and detect overfitting.
:::

::: {.callout-note}
## Prerequisites

This cookbook assumes familiarity with the core `pgdcm` workflow. If you are new to the package, start with the [Beginner Tutorial](https://coinslab.github.io/pgdcm/articles/Beginner_Tutorial.html). For details on constructing attribute hierarchies and higher-order models, see the [Model Cookbook](https://coinslab.github.io/pgdcm/articles/Cookbook.html).
:::

In operational psychometrics, assessment programs rarely estimate everything from scratch each time they administer a test. Doing so is computationally expensive and requires massive sample sizes to guarantee stability.

Instead, testing programs rely on a **Calibration and Scoring** workflow:

1.  **Calibration (Year 1):** Administer the test to a large, representative sample. Run full estimation to establish item parameters (difficulties and discriminations) and validate the underlying cognitive structure.
2.  **Scoring (Year 2+):** Lock those item parameters in place. When new examinees take the test, the model bypasses re-estimation of the test blueprint and focuses exclusively on classifying each examinee's latent masteries against the fixed framework.

This cookbook demonstrates how `pgdcm` abstracts this operational pipeline into a reproducible, automated process.

## Preparing the Data

To simulate a Year 1 vs. Year 2 operational scenario, we split the full 990-examinee `dtmr_data` matrix into two halves: a **Calibration** sample and a **Scoring** sample.

Because we are running a cross-validation experiment later in this document, we also match each examinee's ID to the ground-truth profiles in `dtmr_true_profiles` so we can evaluate classification accuracy at the end.

```{r setup_data}
#| eval: false

library(dcmdata)
library(nimble)
library(pgdcm)

set.seed(2026)
N <- nrow(dtmr_data)
shuffled_idx <- sample(1:N)

# Shuffle data
X_shuffled <- dtmr_data[shuffled_idx, ]

# Match the true profiles to the shuffled IDs
matched_idx <- match(
    as.character(X_shuffled[[1]]),
    as.character(dtmr_true_profiles[[1]])
)
true_profiles_shuffled <- dtmr_true_profiles[matched_idx, ]

# Use string IDs as row names so MCMC output is labeled correctly
rownames(X_shuffled) <- as.character(X_shuffled[[1]])
rownames(true_profiles_shuffled) <- as.character(true_profiles_shuffled[[1]])

# Split into Calibration (Year 1) and Scoring (Year 2) samples
half <- floor(N / 2)
X_calib <- X_shuffled[1:half, ]
X_score <- X_shuffled[(half + 1):N, ]

true_calib <- true_profiles_shuffled[1:half, ]
true_score <- true_profiles_shuffled[(half + 1):N, ]
```

## The Calibration Phase (Year 1)

With the isolated `X_calib` sample, we construct the Q-Matrix graph and run full estimation exactly as in the standard tutorials. For this example we use the **Attribute Hierarchy model (AH-DCM)**:

```{r calibration_ahdcm}
#| eval: false

# Read the pre-specified graphical network
g_ah <- read_graph("AH_DCM.graphml", format = "graphml")

# Compile the model configuration
config_calib <- build_model_config(g_ah, X_calib)

# Estimation settings
estimation_cfg <- list(niter = 10000, nburnin = 1000, chains = 2)

# Run full estimation on the Year 1 sample
res_calib <- run_pgdcm_auto(
    config = config_calib,
    prefix = "scoring_results/AH-DCM_Calib",
    estimation_config = estimation_cfg
)
```

The returned `res_calib` object now contains the posterior distributions for:

-   **`lambda`** — item intercepts and main-effect discriminations.
-   **`theta`** — structural dependency parameters governing attribute relationships.

These posteriors represent the calibrated "blueprint" of the assessment.

## The Scoring Phase (Year 2)

When Year 2 arrives, we need to transfer the calibrated item parameters to score new examinees. Doing this manually in NIMBLE would require subsetting posterior arrays, tracking `beta_root` structural matrices across model types, and injecting tightly constrained priors (`sd = 0.0001`) into initial values.

`build_scoring_config()` handles all of that automatically:

```{r scoring_ahdcm}
#| eval: false

# Extract posteriors from calibration, freeze item parameters,
# and create a scoring configuration targeting the new dataset
config_score <- build_scoring_config(
    calib_results = res_calib,
    calib_config  = config_calib,
    new_dataframe = X_score
)

# Run scoring — the sampler now estimates only the latent profiles
res_score <- run_pgdcm_auto(
    config = config_score,
    prefix = "scoring_results/AH-DCM_Score",
    estimation_config = estimation_cfg
)

# Year 2 diagnostics are ready
head(res_score$skill_profiles)
```

::: {.callout-note}
## What `build_scoring_config()` Does Under the Hood

1.  Extracts the posterior means for `lambda` and `theta` from `res_calib$mapped_parameters`.
2.  Constructs priors with near-zero variance (`sd = 0.0001`) around those means, effectively locking the item parameters.
3.  Rebuilds the full model configuration targeting `X_score`, so the MCMC sampler estimates **only** the latent mastery profiles for the new examinees.
:::


## Cross-Validation Experiment: Detecting Overfitting {#cross-validation}

In a real operational setting, you cannot compute true classification accuracy because the examinees' actual cognitive states are unknown. You must trust that the locked parameters from Year 1 generalize.

Because `dtmr_data` is a simulated dataset, we have the luxury of ground-truth profiles. We can measure exactly how well each model holds up when transferred from Calibration to Scoring. This split-half design exposes the hidden danger of **overfitting**.

### Evaluating Accuracy with `assess_classification_accuracy()`

After each phase, we compare the estimated skill profiles against the known truth:

```{r accuracy_check}
#| eval: false

# Define the mapping between model skill names and ground-truth columns
skills_map <- list(
    "referent_units"           = "referent_units",
    "partitioning_iterating"   = "partitioning_iterating",
    "appropriateness"          = "appropriateness",
    "multiplicative_comparison" = "multiplicative_comparison"
)

# Evaluate the scoring phase accuracy
acc_score <- assess_classification_accuracy(
    skill_profiles = res_score$skill_profiles,
    true_data      = true_score,
    mapping_list   = skills_map
)

acc_score$profile_accuracy
```

### Model Comparison Results

Running this workflow uniformly across DCM, AH-DCM, and HO-DCM produces the following diagnostic table:

| Model | Phase | WAIC | Profile Accuracy |
|:------|:------|-----:|-----------------:|
| DCM | Calibration | 14730.62 | 52.5% |
| DCM | Scoring | 14857.72 | 50.9% |
| AH-DCM | Calibration | 14564.05 | 58.9% |
| **AH-DCM** | **Scoring** | **15079.14** | **57.9%** |
| HO-DCM | Calibration | 14560.84 | 62.0% |
| HO-DCM | Scoring | 15206.87 | 44.6% |

### Interpreting the Results

::: {.callout-warning}
## The Overfitting Trap

Looking only at **Calibration** accuracy, one might conclude that the **HO-DCM** is the best model — it achieves the highest accuracy (62%) and the lowest WAIC (14561). However, its **Scoring** accuracy collapses to 44.6%, the worst of all three models.
:::

This degradation is a textbook example of overfitting. With only $N = 495$ examinees, the HO-DCM's heavily nested hierarchy does not have enough data to reliably estimate 27 items alongside the structural parameters. The model "memorizes" noise in the calibration sample and fails to generalize.

In contrast, the **AH-DCM** loses only ~1 percentage point when moving from Calibration (58.9%) to Scoring (57.9%). Its simpler dependency structure is well-supported by the available sample size, making it the most robust choice for operational deployment.

## Summary

| Step | Function | Purpose |
|:-----|:---------|:--------|
| 1 | `build_model_config()` | Compile the calibration configuration from graph + data |
| 2 | `run_pgdcm_auto()` | Run full MCMC estimation on the calibration sample |
| 3 | `build_scoring_config()` | Lock item parameters and target a new dataset |
| 4 | `run_pgdcm_auto()` | Score new examinees against the frozen framework |
| 5 | `assess_classification_accuracy()` | Evaluate classification against ground truth (if available) |

**Key takeaway:** The `build_scoring_config()` pipeline not only automates operational scoring — it also provides a principled framework for cross-validating model selection. Models that overfit during calibration will reveal themselves through accuracy degradation at scoring time. For the DTMR dataset with $N < 500$, the AH-DCM is the safest operational foundation.
